---
title: "Getting started with birt"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with birt}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

`birt` fits item response theory models with one or two latent variables from
lavaan-style syntax. Typical use: a joint model of response accuracy (binary items, logit
link) and log response times, with ability and speed; or a Likert scale moderated by a
total response time.

By default `birt(model, data)` fits the model by **ELGM**, approximate Bayesian inference
for extended latent Gaussian models (Stringer, Brown & Stafford, 2023), in 'RTMB':

* the person latents are integrated out of the likelihood by per-person adaptive
  Gauss-Hermite quadrature (AGHQ), giving the marginal likelihood;
* the posterior of the remaining parameters, with the priors of `dpriors()`, is integrated
  by an outer adaptive quadrature over the hyperparameters (Gaussian in the other
  directions);
* the output has posterior means, SDs, 95% intervals and the log marginal likelihood, so
  `compare()` gives Bayes factors.

The same marginal likelihood can be maximized instead (`rtmb_control(method = "aghq")`):
maximum likelihood, needed by `score_test()`, `quantile_score()` and `dif_tree()`, and the
natural choice for AIC / BIC. The JAGS engine (`engine = "jags"`, optional) samples the
conditional likelihood and is needed for spike-and-slab priors, `ald()` quantile models,
DIC / WAIC / PSIS-LOO and `ppc()`.

## 1. Model and fit (ELGM)

```{r}
library(birt)
d <- sim_rtirt(N = 500, K = 10)        # y1..y10 (0/1), t1..t10 (log RT), x1, x2
items <- paste0("y", 1:10); times <- paste0("t", 1:10)

model <- rtirt_syntax(items, times)    # van der Linden (2007)
cat(model)                             # ability =~ y's; speed =~ -1*t's; ability ~~ speed

fit <- birt(model, data = d)           # ELGM (default), about 20 s
fit                                    # log marginal likelihood, inner / outer quadrature
summary(fit)                           # posterior means, SDs, 95% intervals, priors
```

* The first `=~` line of a factor gives its measurement loadings; later lines add
  cross-loadings (e.g. `rtirt_syntax(items, times, cross = "free")`: ability on each log RT,
  Bolsinova & Tijmstra, 2018).
* `ability` has variance 1 (it has free loadings); `speed`, whose loadings are all fixed,
  has a free variance; they are uncorrelated unless `ability ~~ speed` is added.
* Priors: defaults from `dpriors()` (`dp = dpriors(cross = "dnorm(0, 100)")` changes a
  class), or `prior("dnorm(0, 100)")*t1` for one parameter (JAGS precision
  parameterization).

## 2. What the object holds

```{r}
estimates(fit)          # lavaan-like table: est, sd, q025, q975, prior
scores(fit)             # EAP and posterior SD per person and latent variable
reliability(fit)        # var(EAP) / (var(EAP) + mean PSD^2)
summary(fit)$items      # a, b (IRT parameterization), RT intercepts, loadings, variances
fit_indices(fit)        # log marginal likelihood
convergence(fit)        # optimizer, outer nodes, quadrature check
equations(fit)          # the model as equations (format = "latex" for a paper)
algorithm(fit)          # how the fit was computed, step by step
```

## 3. Model comparison

```{r}
fc <- birt(rtirt_syntax(items, times, cross = "free"), d)      # item cross-loadings
fi <- birt(sub("ability ~~ speed", "", model, fixed = TRUE), d)  # independent
compare(correlated = fit, cross = fc, independent = fi)          # logML, Bayes factors

# maximum likelihood: AIC and BIC
ml  <- birt(model, d, control = rtmb_control(method = "aghq"))
mlc <- birt(rtirt_syntax(items, times, cross = "free"), d, control = rtmb_control(method = "aghq"))
compare(correlated = ml, cross = mlc)
```

Marginal likelihoods depend on the priors; report them with the priors.

## 4. Distributional models and quantiles

Instead of normal residuals, `shash()` gives sinh-arcsinh residuals (skewness, tail
weight); `V(y) ~ z` in the syntax moderates a log scale. `quantiles(fit, p)` then gives the
effects on every conditional quantile from one fit, with simultaneous bands over `p`;
`quantile_test()` tests whether an effect changes with `p`.

```{r}
reg <- rtirt_syntax(items, times, structure = "path", cov = c("x1", "x2"))  # speed ~ ability + x1 + x2
fs <- birt(reg, d, family = list(speed = shash()), control = rtmb_control(k_hyper = 13))

m5 <- paste(rtirt_syntax(items, times, cross = "free"),
            sprintf("V(%s) ~ ability", paste(times, collapse = " + ")), sep = "\n")
f5 <- birt(m5, d)
quantiles(f5, p = c(0.1, 0.5, 0.9))
quantile_test(f5, p = c(0.1, 0.5, 0.9))
cond_reliability(f5)    # reliability of ability given its value: accuracy, RT, both
rt_information(f5)      # where the RT information lies; what a time limit keeps
```

If ELGM warns that the posterior is too wide for the outer grid along a hyperparameter
(more mass on its two outermost nodes than a Gaussian posterior 1.5 times wider would put
there, and at least 2.5%), check whether that parameter is weakly determined and increase
`rtmb_control(k_hyper = )`.

## 5. Maximum likelihood tools

```{r}
score_test(ml, d$x1 > 0, by_item = TRUE)   # measurement invariance from one fit
score_test(ml, rowSums(d[times]))          # along the total log time
qs <- quantile_score(birt(m5, d, control = rtmb_control(method = "aghq")), p = 0.9)
qs                                         # influence functions, robust SEs
score_test(qs, d$x2)
rtmb_code(ml)                              # a stand-alone RTMB script
```

`dif_tree(model, data, partition)` (package partykit) grows a DIF tree with these tests.

## 6. Ordinal indicators moderated by a response time

With `ordered = TRUE` the indicators are ordinal (graded response model) and person-level
moderators enter as moderation statements on conditional moments, as in MNLFA:
`E(y1 + y2) ~ z` shifts the thresholds (uniform DIF), `E(y1) ~ z:f` moderates the
discrimination (nonuniform DIF), `E(f) ~ z` and `V(f) ~ z` the latent mean and variance.
`hgrm()` writes this syntax from a few options; both fit by ELGM by default.

```{r}
dh <- sim_hgrm(N = 600, J = 8)
dh$logT <- as.numeric(scale(dh$logT))
m <- 'SA =~ y1 + y2 + y3 + y4 + y5 + y6 + y7 + y8
      E(y1 + y2 + y3 + y4 + y5 + y6 + y7 + y8) ~ pa*logT:SA
      E(y1 + y2) ~ logT
      E(SA) ~ logT'
f_ord <- birt(m, dh, ordered = TRUE)
summary(f_ord)
ordinal_information(f_ord)

h0 <- hgrm("SA =~ y1 + y2 + y3 + y4 + y5 + y6 + y7 + y8", dh, moderators = "logT",
           control = rtmb_control(method = "aghq"))
score_test(h0, dh$logT, by_item = TRUE)    # DIF screen, item by item
```

## 7. The JAGS engine (optional)

```{r}
fit_ssp <- birt(rtirt_syntax(items, times, cross = "ssp"), d, engine = "jags")
summary(fit_ssp)$selection    # spike-and-slab: inclusion probabilities, Bayes factors
compare(M2b = fit_ssp, M1 = birt(model, d, engine = "jags"))   # LOOIC, ELPD
ppc(fit_ssp)                  # posterior predictive checks incl. conditional dependence
jags_code(fit_ssp)            # the JAGS model
q <- birt(reg, d, family = list(speed = ald(c(0.1, 0.5, 0.9))), engine = "jags")
```

## 8. Notes

* Covariates are used as given (standardize them) and may not be missing; indicators may
  be missing at random.
* The log-likelihood of response times is that of the log times (no Jacobian), the same
  for `log_rt = TRUE` and for data that are already logged; do not compare it with models
  for times in seconds.
* With ALD indicators the JAGS fit indices are slow; use `fit_indices = FALSE` and
  `add_fit_indices(fit, n_draws = 200)` later. Posterior intervals under ALD working
  likelihoods can be too narrow (Yang, Wang & He, 2016).

## References

Bolsinova, M., & Tijmstra, J. (2018). Improving precision of ability estimation: Getting
more from response times. *British Journal of Mathematical and Statistical Psychology, 71*,
13-38.

Stringer, A., Brown, P., & Stafford, J. (2023). Fast, scalable approximations to posterior
distributions in extended latent Gaussian models. *Journal of Computational and Graphical
Statistics, 32*, 84-98. https://doi.org/10.1080/10618600.2022.2099403

van der Linden, W. J. (2007). A hierarchical framework for modeling speed and accuracy on
test items. *Psychometrika, 72*, 287-308.

Yang, Y., Wang, H. J., & He, X. (2016). Posterior inference in Bayesian quantile regression
with asymmetric Laplace likelihood. *International Statistical Review, 84*, 327-344.
